One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2025-01-08 paper.tex
change author order
most of Janine comments

@@

% seems like this document class uses short author from the last author
% macro in the preamble?  
\author[Andrew E. Seaton \textit{et al.}]{Steve J. Kendall}
\address{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}

@@

The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which are crucial to guide effective conservation management decisions.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution through time.

@@

The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to the island of \hawaii{}, USA.  Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}. Information from the HFBS is used to update the International Union for Conservation of Nature redlist \citep{iucn_redlist_2022}, where the \akepa{} is currently listed as endangered. HFBS data is also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{} (hereafter Hakalau).  Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.

@@

\begin{figure}[!htb]
	\centering
	\includegraphics{figures/study_area_design.pdf}
	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{}.  The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
	\subcaption*{Alt text: A map of the Hakalau study region with the locations of point transects.  The point transects are arranged approximately evenly spaced along straight lines moving roughly east to west.  The southern portion has higher survey effort comopared to the northern portion of the study area.}
	\label{fig:2002studyareapointspt}
\end{figure}

@@

Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled using a parametric function and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.  Under this approach, animal density is uniform within each strata of the survey.

@@

Model-based distance sampling has most commonly been implemented in a two-stage modelling framework.  In the first stage, detectability is estimated using binned counts of observed distances to fit a detection function.  In the second stage, detectability estimates from the first stage are used as an offset term in a generalized additive model framework.  Detections are are also binned into counts within user-defined sampling units (point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.  The \texttt{dsm} package \citep{miller_spatial_2013}, for example, provides tools to do this using the \texttt{R} package \texttt{mgcv} \citep{wood_gam_2017} to implement the spatially-explicit count model.  This two-stage model is known as a \textit{density surface model} and \cite{miller_spatial_2013} provide a review.

@@

A downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, which is a critical assumption of detection functions \citep{buckland_distance_2015}. The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful. These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA for spatial modelling. Here, we present an approach that is inspired by \citet{yuan_point_2017} but with some key differences. We also fit the model using INLA and take advantage of the benefits of the increased computational efficiency compared to MCMC.  However, we use a well-known parametric family of detection functions, that is familiar to users of classical distance sampling methods. This results in components of the model predictor that are non-linear in their parameters, making model fitting infeasible using standard INLA methods \citep{rue_approximate_2009}.  We address this by using a novel method of iterated model fits based on a first-order Taylor expansion of the non-linear model components.  Our approach is implemented in the \texttt{inlabru} package \citep{lindgren_inlabru_2024, bachl_inlabru_2019}, which also has support for spatial data objects.

@@

Distance sampling models specify the thinning function $g(\bs)$ as a function that decays as distance increases.  The type of distance measured depends on the type of survey.  For line transects the perpendicular distance to the transect line is used whereas for point transects it is the horizontal distance to the observer.  For the remainder of the paper we assume a point transect survey design and hence horizontal distance to an observer located at the centre of each point transect.  The thinning probability function is specified as a parametric family of functions. We denote the distance of a point at $\bs$ from an observer as $r(\bs)$.  The \textit{half-normal} detection function is $g(\bs | \sigma) = \exp(-r(\bs)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.  Other types of detection function, such as hazard-rate, negative exponential, can also be used. Their parameters themselves can be modelled using appropriate link functions to incorporate covariates that may affect detectability.  In addition to these parametric forms, additional adjustments have been proposed to increase the flexibility of such models by including trigonometric series expansions \citep{buckland_distance_2015}.  For simplicity, $\sigma$ is modelled using only an intercept parameter, but our approach can readily be extended to include covariates.

@@

To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.  In a slight abuse of notation, $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes a location within the $k$-th point transect. The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$. The integral in the likelihood can be simplified using an assumption that $\lambda(\bs)$ is linear within each sampling unit.  This implies that $\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi)) = 2\lambda(\bs_k)$.  Therefore
\begin{align}
\label{eq-linear-intensity}
	\int_0^{2\pi} \lambda(\bs_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \{\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi) \} g_k(r)\mathrm{d}\theta \nonumber \\
	&= 2\pi \lambda(\bs_k)g_k(r).
\end{align}
We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the spatially-structured random effect described in Section \ref{sec-gmrf}.

@@

To simplify notation below we gather all integration weights into the vectors $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.  Similarly, let $\tl_k = (\tl(\bs_{k1}), \ldots, \tl(\bs_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$, the thinned intensity evaluated at each integration location, and $\tl_{obs} = (\tl(\bs_1), \ldots, \tl(\bs_n))^\intercal$, the thinned intensity evaluated at each observation location.  Then the approximate log-likelihood can be written as
\begin{equation}
\label{approx-log-likelihood}
	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}.
\end{equation}
This approximate likelihood can be expressed as a modified Poisson likelihood by letting $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and constructing a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.  Then the approximate likelihood is
\begin{equation}
\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i),
\end{equation}
where $C$ is a constant.  This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.

The approach taken here is similar to the Berman-Turner device \citep{baddeley_practical_2000, berman_approximating_1992} that is used to fit a wide variety of point process models using GLM software.  The main difference with our approach is that we do not use the assumption that the locations of observed points form a part of the quadrature scheme.

@@

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.  However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.  It is not clear what impact this has on the CV values and this makes the map hard to interpret, even for statisticians who understand the model well. The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values that are also intended to indicate uncertainty.  However, a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  Whether these relative differences matter is dependent on the overall context of the analysis.  Some levels of variability may be acceptable in some contexts and unacceptable in others.

@@

A solution to the problem of interpreting quantile and exceedance probability maps is to use methods that explicitly consider the joint probability of events across all prediction locations. One option is to use excursions sets and excursion functions \citep{bolin_excursion_2015}.  This requires users to choose thresholds of interest (e.g. the definition of a hotspot), and to select an acceptable level of uncertainty in knowing whether the field exceeds this threshold.  This forces us to consider this explicitly and leaves us less at the whim of an arbitrary colour scale.  Crucially, excursion sets and functions can be estimated using Monte Carlo samples and so fit well within our one-stage Bayesian approach.

>>>>

% seems like this document class uses short author from the last author
% macro in the preamble?  
\author{Steve J. Kendall}
\address{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}

@@

@@

The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}. This paper is motivated by the need to analyse wildlife survey data to achieve this aim.  Such data necessarily has spatial and temporal dependencies, as well as complex observation processes, that need to be accounted for in a statistical analysis.  In particular, we present analysis of data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which are crucial to guide effective conservation management decisions.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution through time.

@@

The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to the island of \hawaii{}, USA.  Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}. Information from the HFBS is used to update the International Union for Conservation of Nature redlist \citep{iucn_redlist_2022}, where the \akepa{} is currently listed as endangered. HFBS data are also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{} (hereafter Hakalau).  Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.

@@

\begin{figure}[!htb]
	\centering
	\includegraphics{figures/study_area_design.pdf}
	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{}.  The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
	\subcaption*{Alt text: A map of the Hakalau study region with the locations of point transects.  The point transects are arranged approximately evenly spaced along straight lines moving roughly east to west.  The southern portion has higher survey effort compared to the northern portion of the study area.}
	\label{fig:2002studyareapointspt}
\end{figure}

@@

Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled using a parametric function and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.  Under this approach, animal density is uniform within each stratum of the survey.

@@

Model-based distance sampling has most commonly been implemented in a two-stage modelling framework.  In the first stage, detectability is estimated using binned counts of observed distances to fit a detection function.  In the second stage, detectability estimates from the first stage are used as an offset term in a generalized additive model framework.  Detections are also binned into counts within user-defined sampling units (point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.  The \texttt{dsm} package \citep{miller_spatial_2013}, for example, provides tools to do this using the \texttt{R} package \texttt{mgcv} \citep{wood_gam_2017} to implement the spatially-explicit count model.  This two-stage model is known as a \textit{density surface model} and \cite{miller_spatial_2013} provide a review.

@@

A downside to this approach is that the solutions to the SPDE are not necessarily monotonically decreasing, which is a critical assumption of detection functions \citep{buckland_distance_2015}. The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful. These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA for spatial modelling. Here, we present an approach that is inspired by \citet{yuan_point_2017} but with some key differences. We also fit the model using INLA and take advantage of the benefits of the increased computational efficiency compared to MCMC.  However, we use a well-known parametric family of detection functions, that is familiar to users of classical distance sampling methods. This results in components of the model predictor that are non-linear in their parameters, making model fitting infeasible using standard INLA methods \citep{rue_approximate_2009}.  We address this by using a novel method of iterated model fits based on a first-order Taylor expansion of the non-linear model components.  Our approach is implemented in the \texttt{inlabru} package \citep{lindgren_inlabru_2024, bachl_inlabru_2019}, which also has support for spatial data objects.

@@

Distance sampling models specify the thinning function $g(\bs)$ as a function that decays as distance increases.  The distance measured depends on the type of survey transect.  For line transects the perpendicular distance to the transect line is used, whereas for point transects the distance refers to the distance to the observer at a central, unmoving location.  For the remainder of the paper we assume a point transect survey design and hence horizontal distance to an observer located at the centre of each point transect.  The thinning probability function is specified as a parametric family of functions. We denote the distance of a point at $\bs$ from an observer as $r(\bs)$.  The \textit{half-normal} detection function is $g(\bs | \sigma) = \exp(-r(\bs)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.  Other types of detection function, such as hazard-rate, negative exponential, can also be used. Their parameters themselves can be modelled using appropriate link functions to incorporate covariates that may affect detectability.  In addition to these parametric forms, additional adjustments have been proposed to increase the flexibility of such models by including trigonometric series expansions \citep{buckland_distance_2015}.  For simplicity, $\sigma$ is modelled using only an intercept parameter, but our approach can readily be extended to include covariates.

@@

To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.  In a slight abuse of notation, $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes a location within the $k$-th point transect. The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$. The integral in the likelihood can be simplified using an assumption that $\lambda(\bs)$ is linear within each sampling unit.  This implies that $\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi)) = 2\lambda(\bs_k)$.  Therefore
\begin{align}
\label{eq-linear-intensity}
	\int_0^{2\pi} \lambda(\bs_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \{\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi) \} g_k(r)\mathrm{d}\theta \nonumber \\
	&= 2\pi \lambda(\bs_k)g_k(r).
\end{align}
As mentioned above, this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the spatially-structured random effect described in Section \ref{sec-gmrf}.

@@

To simplify notation below we gather all integration weights into the vectors $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.  Similarly, let $\tl_k = (\tl(\bs_{k1}), \ldots, \tl(\bs_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$, the thinned intensity evaluated at each integration location, and $\tl_{obs} = (\tl(\bs_1), \ldots, \tl(\bs_n))^\intercal$, the thinned intensity evaluated at each observation location.  Then the approximate log-likelihood can be written as
\begin{equation}
\label{approx-log-likelihood}
	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}.
\end{equation}
This approximate likelihood can be expressed as a modified Poisson likelihood by letting $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and constructing a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.  Then the approximate likelihood is
\begin{equation}
\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i),
\end{equation}
where $C$ is a constant.

The approach taken here is similar to the Berman-Turner device \citep{baddeley_practical_2000, berman_approximating_1992} that is used to fit a wide variety of point process models using GLM-type fitting software.  The main difference with our approach is that we do not use the assumption that the locations of observed points form a part of the quadrature scheme. This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.

@@

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.  However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.  It is not clear what impact this has on the CV values and this makes the map hard to interpret, even for statisticians who understand the model well. The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values that are also intended to indicate uncertainty.

While these measures of uncertainty do convey some information, our main issue with plots like these is that a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  Whether these relative differences matter is a separate question, one that is dependent on the overall context of the analysis.  Some levels of variability may be acceptable in some contexts and unacceptable in others.  It is often unclear what these uncertainty maps actually mean to us in the wider context of an analysis.  Below we advocate for an approache that requires more thought about what levels of uncertainty are acceptable for a given aim, and to produce model outputs that incorporate this thinking.  The Bayesian one-stage approach is well suited to support this kind of bespoke model output.

@@

A solution to the problem of interpreting quantile and exceedance probability maps is to use methods that explicitly consider the joint probability of events across all prediction locations. One option is to use excursions sets and excursion functions \citep{bolin_excursion_2015}.  These approaches require users to choose thresholds of interest (e.g. the definition of a hotspot), and to select an acceptable level of uncertainty in knowing whether the field exceeds this threshold.  This is in contrast to the issues we mentioned above regarding default colour scales and unclear meanings of uncertainty maps.  Crucially, excursion sets and functions can be estimated using Monte Carlo samples and so fit well within the one-stage Bayesian approach.
